How environmental drivers of spatial synchrony interact

### 2023-05-30 Paper/Paper.Rnw
version actually submitted to Ecography
changes in response to referees, minor revisions

%external documents
\externaldocument[SM-]{SupMat}

@@

%\doublespacing

@@

\begin{abstract}
\noindent Spatial synchrony, the tendency for populations across space to show correlated fluctuations, is a fundamental 
feature of population dynamics, linked to central topics of ecology such as population cycling, extinction 
risk, and ecosystem stability. 
A common mechanism of spatial synchrony is the Moran effect, whereby spatially synchronized environmental signals 
drive population dynamics and hence induce population synchrony. 
%***DAN: the below line is the new line added to fit into Ecography, delete if you don't go that way
After reviewing recent progress in understanding Moran effects, 
%***DAN: the above line is the new line added to fit into Ecography, delete if you don't go that way
we here elaborate a general theory of how 
Moran effects of different environmental drivers acting on the same populations can interact, either 
synergistically or destructively, to produce either substantially more or markedly less population synchrony 
than would otherwise occur. 
We provide intuition for how this newly recognized mechanism works through theoretical case studies and 
application of our theory to California populations of giant kelp.
We argue that Moran interactions should be common.     
Our theory and analysis explain an important new aspect of a fundamental feature of spatiotemporal population dynamics. 
\end{abstract}

\keywords{Fourier, interactions, kelp, \emph{Macrocystis pyrifera}, Moran effects, nutrients, synchrony, waves}

@@

%Intro to synchrony
Spatial synchrony, the tendency for geographically disjunct populations to show correlated fluctuations 
through time, is a fundamental feature of population dynamics, linked to important topics such as 
population cycling \citep{Anderson2021}, extinction risk \citep{Ghosh2020}, and the 
stability of regional populations and ecosystem functioning \citep{Wilcox2017}. Though spatial synchrony (henceforth, 
\emph{synchrony}) has been studied for decades
in a wide variety of species ranging from viruses and plants to mammals, and at 
spatial scales ranging from centimeters to over 1000 km \citep{Liebhold2004},
recent advances in statistical methods and improvements in data availability have led to several 
major advances in our understanding of synchrony and its causes and consequences. 
For instance, the timescale structure of synchrony is now known to be 
important [\cite{Sheppard2016,Desharnais2018,Sheppard2019}, see also \cite{Keitt2008}],
synchrony is now known to have a complex and pronounced geography \citep{Defriez2017,Defriez2017b,Walter2017,Walter2020}, 
and patterns of synchrony are now known to be asymmetric in distribution tails in a way that
is important for system stability \citep{Ghosh2020b,Ghosh2020c,Walter2021}. Our ability to infer the causes of 
synchrony is much improved in recent years \citep{Sheppard2016,Defriez2017b,Walter2017,Anderson2018,Sheppard2019}, 
and there is also growing evidence
that synchrony is changing as a newly recognized component of climate 
change \citep{Defriez2016,Sheppard2016,Hansen2020}. 
Distinct viewpoints on synchrony from population and community ecology are 
becoming integrated, leading to a more wholistic understanding 
of the importance of synchrony for ecosystem stability \citep{Wang2014}. These new developments 
have led to an increasingly central role of the phenomenon of synchrony in many of the most 
important research topics in ecology.

%Intro to Moran effects and history of the early years where people had trouble identifying them
%***DAN: new review para for ecography
Correlations between weather time series measured in different locations can 
induce synchrony between populations in those locations if the weather variables influence
population processes. This mechanism, called the Moran effect, is now known to be one of the most
important causes of synchrony. But the mechanism was originally proposed 
theoretically \citep{Moran1953}, and decades passed during which it was 
considered difficult to distinguish this potential cause of synchrony from others in field systems.
Synchrony has long been thought to have three causes: dispersal, and trophic interactions with a synchronous or
mobile species, in addition to the Moran effect \citep{Liebhold2004}. However, using common past statistical approaches
which focussed on declines in population correlations with distance, patterns of synchrony produced
by each of these mechanisms are quite similar \citep{Ranta1999,Abbott2007,Liebhold2004,Walter2017}; so examination
of such patterns provides little or no traction for inferring the causes of synchrony. 
Early papers demonstrating the Moran effect mechanism resorted to special cases where dispersal was 
impossible and predators were absent \citep{Grenfell1998,Tedesco2004}. Controlled experiments also confirmed that all
three theorized mechanisms could be involved in synchrony [e.g., \citealp{Vasseur2009}]. Nevertheless, the problem of inferring
specific mechanisms of synchrony in field systems was considered a challenge 
until recently.

%Recent statistical explorations of Moran effects
%***DAN: new review para for ecography
Recent research has provided new statistical viewpoints which have, when sufficient data are available, 
essentially solved the problem of inference
of the causes of synchrony, and the research has revealed the broad importance of Moran effects. 
%geographic approaches
Approaches based on formal statistical comparisons between population synchrony maps and maps of 
environmental correlations produced inferences that precipitation and temperature Moran effects are important
causes of synchrony of terrestrial \citep{Defriez2017b} and marine \citep{Defriez2017} primary productivity. 
Another geographic approach, based on multiple regression of distance matrices (MRM), was used in an early paper
to infer that a precipitation Moran effect is a cause of synchrony for the spongy moth \citep{Haynes2013},
an invasive forest pest whose name was recently changed from ``gypsy moth.'' 
Geographic approaches to identifying causes of synchrony, often MRM approaches 
specifically \citep{Koenig2017,Walter2020,Bogdziewicz2021}, 
have become widespread, and the geography of synchrony has become a mainstream area of study \citep{Walter2017}.
The approach of \cite{Gouveia2016}, called ``noncentered local indicators of spatial association'' (ncLISA), is another 
geographic approach; as is the ``fluvial synchrogram'' of \cite{Larsen2021}, a tool for studying synchrony in 
river networks.
MRM approaches have also been used to identify dispersal as a cause of synchrony \citep{Anderson2018},
sometimes combined with genetic methods \citep{Haynes2022}.
%Time and timescale methods
Another class of methods exploit the time and timescale structure of synchrony.
Using a method that examined the match between temporal variations in population and environmental synchrony, 
\cite{Allstadt2015} confirmed the importance of a precipitation Moran effect for synchrony of spongy moth.
A suite of wavelet and related Fourier techniques have been developed that can comprehensively describe the time and timescale
structure of synchrony, identify Moran drivers of synchrony opperating in distinct timescale bands, and apportion fractions
of observed synchrony to respective Moran drivers \citep{Sheppard2016,Sheppard2017,Desharnais2018,Sheppard2019,wsyn,Anderson2021}.
The methods have been used, for instance, to provide one of the first clear demonstrations that changes in 
Moran effects are another of the impacts of climatic changes \citep{Sheppard2016,Sheppard2019}; and to illustrate the importance of the
timescale structure of environmental covariation for altering Moran effects \citep{Desharnais2018}.
These methods have since been used to identify Moran effects in several additional systems
(e.g., \citealp{Anderson2019,Walter2019,Walter2020,GarciaCarreras2022,CastoraniSheppardKelp}).

@@

%Demo that interactions can happen
The basic fact that Moran drivers can interact, synergistically or destructively, can be demonstrated 
with a very simple model which we now elaborate, though understanding the full nature of the interaction
mechanism will require the rest of this paper. Suppose $\epsilon_i^{(1)}(t)$ and $\epsilon_i^{(2)}(t)$
are environmental random variables measured in locations $i=1,2$ at times $t$, and assume these are independent
through time and standard normally distributed for all $i$ and $t$. 
If a population index $w_i(t)$ 
follows the autoregressive process $w_i(t)=b w_i(t-1)+\epsilon_i^{(1)}(t)$ for $i=1,2$, then the classic Moran theorem 
\citep{Moran1953} implies that temporal population correlation equals temporal environmental correlation, i.e., 
$\cor(w_1,w_2)=\cor(\epsilon_1^{(1)},\epsilon_2^{(1)})$. If, instead, 
$w_i(t)=b w_i(t-1)+\epsilon_i^{(1)}(t)+\epsilon_i^{(2)}(t)$, so that both environmental variables influence 
populations, the Moran theorem then implies that population synchrony is $\cor(w_1,w_2)=
\cor(\epsilon_1^{(1)}+\epsilon_1^{(2)},\epsilon_2^{(1)}+\epsilon_2^{(2)})$.
As we show in detail in SI section \ref{SM-Sect:IntroExample}, it is straightforward to show that 
this quantity depends not only on the standard environmental synchrony 
measures $\cov(\epsilon_1^{(a)},\epsilon_2^{(a)})$, for 
$a=1,2$, but also on the ``cross synchrony'' measures $\cov(\epsilon_1^{(a)},\epsilon_2^{(b)})$
for $a \neq b$. 
Cross synchrony measures represent synchrony between an environmental variable in one location and 
a different variable in another location. Fig. \ref{fig:IntroTheory} illustrates, using this model, 
how population synchrony can 
therefore differ from what synchrony would be if the environmental processes $\epsilon^{(1)}$ and 
$\epsilon^{(2)}$ were unrelated, so that covariances between components of those processes are zero. 
The difference is due to interactions
between the two Moran effects, and can be substantially positive or negative. This example was adapted 
from SI section S1 of \cite{Sheppard2019}. All notation and abbreviations used throughout the paper 
are summarized in Table \ref{tab:notation}.

@@

%Final intro para with specific goals
Our specific goals for this study are as follows. First, (G1) we will elaborate a general 
theory of interacting Moran effects which allows a detailed understanding of how the mechanism works. Second, 
(G2) we will consider theoretical case studies, which emerge as important special cases of our general theory, and 
which illuminate the intuition behind how Moran effects may interact in real systems. Third, (G3) we will 
apply our theory to populations of giant kelp off the CA coast. Whereas \cite{CastoraniSheppardKelp}
already carried out a detailed analysis of interacting Moran effects in CA kelp populations and their importance 
for kelp ecology, we instead use a simplified subset of the available kelp data. Our kelp analysis is intended 
to illuminate the inner workings of the general mechanism we describe rather than being an exploration
of kelp dynamics, specifically. Overall, this study introduces a general theory of Moran interactions and uses it to conceptually
illuminate this newly observed but potentially quite important mechanism of spatiotemporal population dynamics.

@@

\noindent Prior to presenting our formal theory, we extend and formalize the intution behind it
that began with the swing analogy in the Introduction. We again focus on a single
timescale of oscillation, later combining timescales mathematically. 
Fig. \ref{fig:PedagogAlt}a,b shows the period-20 Fourier components of two hypothetical,
spatially synchronous environmental variables measured in sampling locations 
$i=1,2,3$, and one way these can influence populations. The components are
lagged, relative to each other, in the timing of their peaks ($l_n$ on the figure). They are also lagged in their 
effects on populations, by the amounts $l_{e1}$ and $l_{e2}$, respectively, i.e., peaks in the environmental signals
manifest as maximum positive influence on populations after delays of $l_{e1}$ and $l_{e2}$, respectively. 
Such delays can be due to a variety of biological mechanisms associated with the life history of the organisms 
which comprise the populations. In this example,
we assume for simplicity that larger values of both environmental variables are beneficial to 
populations, though the general
theory described below does not require that assumption, and see also the next example which makes a
different assumption. If, as in Fig. \ref{fig:PedagogAlt}a, b, $l_{e1}-l_{e2}+l_n=0$ (or $l_{e1}-l_{e2}+l_n$ is any integer
multiple of $\sigma=20$, the period), then the 
periodic maximal positive influences of the two environmental variables coincide with each other as well as being 
spatial synchronous. This alignment of influences produces additional synchrony in populations, 
beyond what would manifest if environmental fluctuations were unrelated, because positive 
influences of variable 1 in location $i$ will
tend to coincide with positive influences of variable 2 in location $j$. In contrast, Fig. \ref{fig:PedagogAlt}c, d
shows the opposite scenario, for which $l_{e1}-l_{e2}+l_n=-\sigma/2$. Thus
the periodic maximal positive influence of environmental variable 1 coincides with the periodic maximal negative
influence of environmental variable 2, reducing synchrony relative to the case of unrelated environmental variables. 
(The same outcome would occur if $l_{e1}-l_{e2}+l_n$ were any integer multiple of $\sigma$ plus or minus $\sigma/2$.)
Intermediate scenarios between the two scenarios of Fig. \ref{fig:PedagogAlt} are also possible, as will be revealed by 
the theory. We henceforth refer to the quantity $l_{e1}-l_{e2}+l_n$ as the \emph{environmental effect alignment measure},
because it measures the extent to which the timing of the population influences of the two environmental variables
are aligned. If we replace the assumption that
larger values of both environmental variables are beneficial to populations with an assuption that larger 
values of the first variable and smaller values of the second are beneficial, scenarios of synergistic versus
antagonistic Moran interactions are reversed, but with the same general concepts still operating 
(Fig. \ref{fig:PedagogAlt2}).

@@

Our formal theory requires a conceptual understanding of the \emph{spectrum} of an environmental 
or population time series, as well as of the \emph{cospectrum} and \emph{cross spectrum} of two time series,
so we briefly introduce these concepts.
If $w_i(t)$ is a stochastic process or time series measured at location $i$ (e.g., the population density time series
of a species of interest in that location), the spectrum $S_{w_i w_i}(f)$ is a function
of frequency, $f$. For a periodic ocillation, $f$ can be defined as one over 
the timescale, or period, $\sigma$, of the oscillation. 
The spectrum $S_{w_i w_i}(f)$ is larger for frequencies at which $w_i(t)$ oscillates more. So, for example, a 
population $w_i(t)$ that shows strong oscillatory dynamics of $5$-year period will have a large value of 
$S_{w_i w_i}(f)$ for $f=\frac{1}{5} \text{ y}^{-1}$. The spectrum separates the overall variance of a time series into
contributions which occur at different frequencies, in the sense that an integral of $S_{w_i w_i}(f)$ across all 
frequencies equals $\var(w_i)$. In a similar way, the cospectrum of two time series, $w_i(t)$ and $w_j(t)$,
is a function of $f$ that takes large values at frequencies for which oscillations in $w_i(t)$ and 
$w_j(t)$ are both strong and strongly correlated, i.e., they have the same or similar phase. 
Here, $j$ refers to another location where measurements were taken. The cospectrum is 
the real part of the cross spectrum, $S_{w_i w_j}(f)$, which is a complex-valued function of frequency.
The cross-spectrum takes large-magnitude values at frequencies, $f$, for which the oscillatory components
of $w_i(t)$ and $w_j(t)$ are strong and in a consistent phase relationship to each other; 
and the complex phase of the cross spectrum at $f$ then quantifies
that relationship. The notation $S_{ww}(f)$ refers to the \emph{spectral matrix}, 
which has $ij^{\text{th}}$ entry $S_{w_i w_j}(f)$. Spectral methods are standard  
[e.g., \cite{Vasseur2007,Defriez2017}], and many background references are available 
[e.g., \cite{Brillinger2001}].

Much prior work demonstrates the importance of a frequency- or timescale-specific approach to
synchrony [e.g., \cite{Vasseur2007,Keitt2008,Defriez2016,Sheppard2016,Desharnais2018,Anderson2021}; frequency- and timescale-specific approaches
are equivalent because frequency and timescale are reciprocal], 
and it will turn out (see below, and Discussion) that a timescale-specific 
approach is essential to the development of our new theory. 
We therefore here define, using the spectral methods outlined above, a frequency/timescale-specific measure of synchrony, 
as well as a new concept of \emph{cross-variable synchrony}. If time series data $w_i(t)$ for $t=1,\ldots,T$ 
were gathered at locations $i=1,\ldots,N$, our synchrony measure is simply
$\rho_{ww}=\left( \sum_{i \neq j} S_{w_i w_j} \right)/(N^2-N)$, the average of the cross spectra for all pairs
of distinct locations. This is a real-valued function of frequency, an integral (across frequencies) of which
is the classic, non-frequency-specific synchrony 
measure $\left( \sum_{i \neq j} \cov(w_i,w_j) \right)/(N^2-N)$ (see SI section
\ref{SM-Sect:measuring_sync} for details, here). If two time series $\epsilon_i^{(1)}(t)$ and $\epsilon_i^{(2)}(t)$
($t=1,\ldots,T$ and $i=1,\ldots,N$) were measured at each sampling location (e.g., two environmental variables),
cross-variable synchrony (or, simply, \emph{cross synchrony}) between the variables
is defined as $\rho_{\epsilon^{(1)}\epsilon^{(2)}}=\left( \sum_{i \neq j} S_{\epsilon_i^{(1)}\epsilon_j^{(2)}} \right)/(N^2-N)$. 
This is interpretable as pertaining to spatial synchrony because it makes comparisons across 
distinct locations, i.e., $i \neq j$. 
It is interpretable as cross-variable synchrony because comparisons are between the two variables, i.e., cross spectra 
between time series components of $\epsilon^{(1)}$ and $\epsilon^{(2)}$ are used. The new index 
takes into account possible time lags. For instance, if both $\epsilon^{(1)}$ and $\epsilon^{(2)}$ show strong,
spatially synchronous, $4$-year-period 
oscillations, but peaks in the $\epsilon^{(2)}$ oscillations consistently lag peaks in the $\epsilon^{(1)}$ oscillations 
by a year, then $\rho_{\epsilon^{(1)}\epsilon^{(2)}}$, which is complex valued, will have high magnitude at
timescale $4$y, equivalent to frequency $f=\frac{1}{4} \text{ y}^{-1}$, and will have  
phase at that frequency equal to $\pi/2$, reflecting the one-year lag.
See SI section \ref{SM-sect:examples_cross_sync} for detailed examples.

>>>>

@@

\doublespacing

@@

\newpage

\large{\textbf{How environmental drivers of spatial synchrony interact}}

\begin{abstract}
\noindent Spatial synchrony, the tendency for populations across space to show correlated fluctuations, is a fundamental 
feature of population dynamics, linked to central topics of ecology such as population cycling, extinction 
risk, and ecosystem stability. 
A common mechanism of spatial synchrony is the Moran effect, whereby spatially synchronized environmental signals 
drive population dynamics and hence induce population synchrony. 
After reviewing recent progress in understanding Moran effects, 
we here elaborate a general theory of how 
Moran effects of different environmental drivers acting on the same populations can interact, either 
synergistically or destructively, to produce either substantially more or markedly less population synchrony 
than would otherwise occur. 
We provide intuition for how this newly recognized mechanism works through theoretical case studies and 
application of our theory to California populations of giant kelp.
We argue that Moran interactions should be common.     
Our theory and analysis explain an important new aspect of a fundamental feature of spatiotemporal population dynamics. 
\end{abstract}

\keywords{Fourier, interactions, kelp, \emph{Macrocystis pyrifera}, Moran effects, synchrony}

@@

%Intro to synchrony
Spatial synchrony, the tendency for geographically disjunct populations to show correlated fluctuations 
through time, is a fundamental feature of population dynamics, linked to important topics such as 
population cycling \citep{Anderson2021}, extinction risk \citep{Ghosh2020}, and the 
stability of regional populations and ecosystem functioning \citep{Wilcox2017}. Though spatial synchrony (henceforth, 
\emph{synchrony}) has been studied for decades
in a wide variety of species ranging from viruses and plants to mammals, and at 
spatial scales ranging from centimeters to over 1000 km \citep{Liebhold2004},
recent advances in statistical methods and improvements in data availability have led to several 
major advances in our understanding of synchrony and its causes and consequences. 
For instance, the timescale structure of synchrony is now known to be 
important [\cite{Sheppard2016,Desharnais2018,Sheppard2019}, see also \cite{Keitt2008}],
synchrony is now known to have a complex and pronounced geography \citep{Defriez2017,Defriez2017b,Walter2017,Walter2020}, 
and patterns of synchrony are now known to be ``asymmetric in distribution tails'' in a way that
is important for system stability \citep{Ghosh2020b,Ghosh2020c,Walter2021}. 
Synchrony is asymmetric in distribution tails when two positively associated (synchronous) variables are 
more (respectively, less) synchronous in the lower parts of their distributions than in the 
upper parts. Our ability to infer the causes of 
synchrony is much improved in recent years \citep{Sheppard2016,Defriez2017b,Walter2017,Anderson2018,Sheppard2019}, 
and there is also growing evidence
that synchrony is changing as a newly recognized component of climate 
change \citep{Defriez2016,Sheppard2016,Hansen2020}. 
Distinct viewpoints on synchrony from population and community ecology are 
becoming integrated, leading to a more wholistic understanding 
of the importance of synchrony for ecosystem stability \citep{Wang2014}. These new developments 
have led to an increasingly central role of the phenomenon of synchrony in many of the most 
important research topics in ecology.

%Intro to Moran effects and history of the early years where people had trouble identifying them
Correlations between weather time series measured in different locations can 
induce synchrony between populations in those locations if the weather variables influence
population processes. This mechanism, called the Moran effect, is now known to be one of the most
important causes of synchrony. But the mechanism was originally proposed 
theoretically \citep{Moran1953}, and decades passed during which it was 
considered difficult to distinguish this potential cause of synchrony from others (e.g., dispersal) in field systems.
Synchrony has long been thought to have three causes: dispersal, and trophic interactions with a synchronous or
mobile species, in addition to the Moran effect \citep{Liebhold2004}. However, using common past statistical approaches
which focussed on declines in population correlations with distance, patterns of synchrony produced
by each of these mechanisms are quite similar \citep{Ranta1999,Abbott2007,Liebhold2004,Walter2017}; so examination
of such patterns provides little or no traction for inferring the causes of synchrony. 
Early empirical papers demonstrating the Moran effect mechanism resorted to special cases where dispersal was 
impossible and predators were absent \citep{Grenfell1998,Tedesco2004}. Controlled experiments also confirmed that all
three theorized mechanisms could be involved in synchrony [e.g., \citealp{Vasseur2009}]. Nevertheless, the problem of inferring
specific mechanisms of synchrony in field systems was considered a challenge 
until recently.

%Recent statistical explorations of Moran effects
Recent research has provided new statistical viewpoints which have, when sufficient data are available, 
essentially solved the problem of inference
of the causes of synchrony (including identifying specific Moran drivers), and the research has 
revealed the broad importance of Moran effects.
%geographic approaches
Approaches based on spatial statistics produced inferences that precipitation and temperature Moran effects are important
causes of synchrony of terrestrial \citep{Defriez2017b} and marine \citep{Defriez2017} primary productivity. 
Another geographic approach, based on multiple regression of distance matrices (MRM), was used 
to infer that a precipitation Moran effect is a cause of synchrony for the spongy moth \citep{Haynes2013}. 
Geographic approaches to identifying causes of synchrony \citep{Koenig2017,Walter2020,Bogdziewicz2021} 
have become widespread \citep{Gouveia2016,Walter2017,Larsen2021}.
MRM approaches have also been used to identify dispersal as a cause of synchrony \citep{Anderson2018},
sometimes combined with genetic methods \citep{Haynes2022}.
%Time and timescale methods
Another class of methods exploit the time and timescale structure of synchrony.
E.g., \cite{Allstadt2015} confirmed the importance of a precipitation Moran effect in spongy moth.
A suite of wavelet and related Fourier techniques have been developed that can comprehensively 
describe the time and timescale
structure of synchrony, identify Moran drivers of synchrony operating in distinct timescale bands, and apportion fractions
of observed synchrony to respective Moran 
drivers \citep{Sheppard2016,Sheppard2017,Desharnais2018,Sheppard2019,wsyn,Anderson2021}.

@@

%Demo that interactions can happen
The basic fact that Moran drivers can interact, synergistically or destructively, can be demonstrated 
with a very simple model which we now elaborate, though understanding the full nature of the interaction
mechanism will require the rest of this paper. Suppose $\epsilon_i^{(1)}(t)$ and $\epsilon_i^{(2)}(t)$
are environmental random variables measured in locations $i=1,2$ at times $t$, and assume these are independent
through time and standard normally distributed for all $i$ and $t$. 
If a population index $w_i(t)$ 
follows the autoregressive process $w_i(t)=b w_i(t-1)+\epsilon_i^{(1)}(t)$ for $i=1,2$, then the classic Moran theorem 
\citep{Moran1953} implies that temporal population correlation equals temporal environmental correlation, i.e., 
$\cor(\epsilon_1^{(1)},\epsilon_2^{(1)})$. If, instead, 
$w_i(t)=b w_i(t-1)+\epsilon_i^{(1)}(t)+\epsilon_i^{(2)}(t)$, so that both environmental variables influence 
populations, the Moran theorem then implies that population synchrony is 
$\cor(\epsilon_1^{(1)}+\epsilon_1^{(2)},\epsilon_2^{(1)}+\epsilon_2^{(2)})$.
As we show in detail Supporting information, it is straightforward to show that 
this quantity depends not only on the standard environmental synchrony 
measures $\cov(\epsilon_1^{(1)},\epsilon_2^{(1)})$ and $\cov(\epsilon_1^{(2)},\epsilon_2^{(2)})$, 
but also on ``cross synchrony'' measures such as $\cov(\epsilon_1^{(1)},\epsilon_2^{(2)})$
and $\cov(\epsilon_1^{(2)},\epsilon_2^{(1)})$. 
Cross synchrony measures represent synchrony between an environmental variable in one location and 
a different variable in another location. Fig. \ref{fig:IntroTheory} illustrates, using this model, 
how population synchrony can 
therefore differ from what synchrony would be if the environmental processes $\epsilon^{(1)}$ and 
$\epsilon^{(2)}$ were unrelated. The difference is due to interactions
between the two Moran effects, and can be substantially positive or negative. This example was adapted 
from SI section S1 of \cite{Sheppard2019}. All notation and abbreviations used throughout the paper 
are summarized in Table \ref{tab:notation}.

@@

%Final intro para with specific goals
Our specific goals for this study are as follows. First, (G1) we will elaborate a general 
theory of interacting Moran effects which allows a detailed understanding of how the mechanism works. Second, 
(G2) we will consider theoretical case studies, which emerge as important special cases of our general theory, and 
which illuminate the intuition behind how Moran effects may interact in real systems. Third, (G3) we will 
apply our theory to populations of giant kelp off the CA coast. Whereas \cite{CastoraniSheppardKelp}
already carried out a detailed analysis of interacting Moran effects in CA kelp populations and their importance 
for kelp ecology, we instead use a simplified subset of the available kelp data. 
Our kelp results are not intended to extend what is known about kelp ecology (beyond what was discovered 
in our earlier paper, \citealp{CastoraniSheppardKelp}), but rather serve as a
vehicle for understanding the inner workings of our general theory of interacting Moran effects.
Overall, this study introduces a general theory of Moran interactions and uses it to conceptually
illuminate this newly observed but potentially quite important mechanism of spatiotemporal population dynamics.

@@

\noindent Prior to presenting our formal theory, we extend and formalize the intution behind it
that began with the swing analogy in the Introduction. We again focus on a single
timescale of oscillation, later combining timescales mathematically. 
Fig. \ref{fig:PedagogAlt}a,b shows the period-20 Fourier components of two hypothetical,
spatially synchronous environmental variables measured in sampling locations 
$i=1,2,3$, and one way these can influence populations. The components are
lagged, relative to each other, in the timing of their peaks ($l_n$ on the figure). They are also lagged in their 
effects on populations, by the amounts $l_{e1}$ and $l_{e2}$, respectively, i.e., peaks in the environmental signals
manifest as maximum positive influence on populations after delays of $l_{e1}$ and $l_{e2}$, respectively. 
Such delays can be due to a variety of biological mechanisms associated with the life history of the organisms 
which comprise the populations. In this example,
we assume for simplicity that larger values of both environmental variables are beneficial to 
populations, though the general
theory described below does not require that assumption, and see also the next example which makes a
different assumption. If, as in Fig. \ref{fig:PedagogAlt}a, b, $l_{e1}-l_{e2}+l_n=0$ (or $l_{e1}-l_{e2}+l_n$ is any integer
multiple of $\sigma=20$, the period), then the 
periodic maximal positive influences of the two environmental variables coincide with each other as well as being 
spatial synchronous. This alignment of influences produces additional synchrony in populations, 
beyond what would manifest if environmental fluctuations were unrelated, because positive 
influences of variable 1 in location $i$ will
tend to coincide with positive influences of variable 2 in location $j$. In contrast, Fig. \ref{fig:PedagogAlt}c, d
shows the opposite scenario, for which $l_{e1}-l_{e2}+l_n=-\sigma/2$. Thus
the periodic maximal positive influence of environmental variable 1 coincides with the periodic maximal negative
influence of environmental variable 2, reducing synchrony relative to the case of unrelated environmental variables. 
(The same outcome would occur if $l_{e1}-l_{e2}+l_n$ were any integer multiple of $\sigma$ plus or minus $\sigma/2$.)
Intermediate scenarios between the two scenarios of Fig. \ref{fig:PedagogAlt} are also possible, as will be revealed by 
our theory. We henceforth refer to the quantity $l_{e1}-l_{e2}+l_n$ as the \emph{environmental effect alignment measure},
because it measures the extent to which the timing of the population influences of the two environmental variables
are aligned. If we replace the assumption that
larger values of both environmental variables are beneficial to populations with an assuption that larger 
values of the first variable and smaller values of the second are beneficial, scenarios of synergistic versus
antagonistic Moran interactions are reversed, but with the same general concepts still operating 
(Supporting information analogue of Fig. \ref{fig:PedagogAlt}).

@@

Our formal theory requires a conceptual understanding of the \emph{spectrum} of an environmental 
or population time series, as well as of the \emph{cospectrum} and \emph{cross spectrum} of two time series,
so we briefly introduce these concepts.
If $w_i(t)$ is a stochastic process or time series measured at location $i$ (e.g., the population density time series
of a species of interest in that location), the spectrum $S_{w_i w_i}(f)$ is a function
of frequency, $f$. For a periodic ocillation, $f$ can be defined as one over 
the timescale, or period, $\sigma$, of the oscillation. 
The spectrum $S_{w_i w_i}(f)$ is larger for frequencies at which $w_i(t)$ oscillates with 
larger amplitude. So, for example, a 
population $w_i(t)$ that shows strong oscillatory dynamics of $5$-year period will have a large value of 
$S_{w_i w_i}(f)$ for $f=\frac{1}{5} \text{ y}^{-1}$. The spectrum separates the overall variance of a time series into
contributions which occur at different frequencies, in the sense that an integral of $S_{w_i w_i}(f)$ across all 
frequencies equals $\var(w_i)$. In a similar way, the cospectrum of two time series, $w_i(t)$ and $w_j(t)$,
is a function of $f$ that takes large values at frequencies for which oscillations in $w_i(t)$ and 
$w_j(t)$ are both strong and strongly correlated, i.e., they have the same or similar phase. 
Here, $j$ refers to another location where measurements were taken. The cospectrum is 
the real part of the cross spectrum, $S_{w_i w_j}(f)$, which is a complex-valued function of frequency.
The cross-spectrum takes large-magnitude values at frequencies, $f$, for which the oscillatory components
of $w_i(t)$ and $w_j(t)$ are strong and in a consistent phase relationship to each other; 
and the complex phase of the cross spectrum at $f$ then quantifies
that relationship. The notation $S_{ww}(f)$ refers to the \emph{spectral matrix}, 
which has $ij^{\text{th}}$ entry $S_{w_i w_j}(f)$. The notation $S_{w_i w_i}$ for the spectrum makes sense
and is consistent with the notation $S_{w_i w_j}$ for the cross spectrum because the spectrum of $w_i$ is the 
cross spectrum of $w_i$ with itself. Spectral methods are standard  
[e.g., \cite{Vasseur2007,Defriez2017}], and many background references are available 
[e.g., \cite{Brillinger2001}].

Much prior work demonstrates the importance of a frequency- or timescale-specific approach to
synchrony [e.g., \cite{Vasseur2007,Keitt2008,Defriez2016,Sheppard2016,Desharnais2018,Anderson2021}; frequency- and timescale-specific approaches
are equivalent because frequency and timescale are reciprocal], 
and it will turn out (see below, and Discussion) that a timescale-specific 
approach is essential to the development of our new theory. 
We therefore here define, using the spectral methods outlined above, a frequency/timescale-specific measure of synchrony, 
as well as a new concept of \emph{cross-variable synchrony}. If time series data $w_i(t)$ for $t=1,\ldots,T$ 
were gathered at locations $i=1,\ldots,N$, our synchrony measure is simply
$\rho_{ww}=\left( \sum_{i \neq j} S_{w_i w_j} \right)/(N^2-N)$, the average of the cross spectra for all pairs
of distinct locations. This is a real-valued function of frequency, an integral (across frequencies) of which
is the classic, non-frequency-specific synchrony 
measure $\left( \sum_{i \neq j} \cov(w_i,w_j) \right)/(N^2-N)$ (see Supporting information 
for details, here, and additional comparisons to previous measures of synchrony). 
Because $\rho_{ww}$ is a real-valued function of frequency, it equals $\rho_{ww}=\Rp(\rho_{ww})=
\left( \sum_{i \neq j} \Rp(S_{w_i w_j}) \right)/(N^2-N)$, which only involves the cospectrum, $\Rp(S_{w_i w_j})$,
not the full cross-spectrum, $S_{w_i w_j}$.
If two time series $\epsilon_i^{(1)}(t)$ and $\epsilon_i^{(2)}(t)$ (e.g., two environmental variables)
were measured at each sampling location and time,
cross-variable synchrony (or, simply, \emph{cross synchrony}) between the variables
is defined as $\rho_{\epsilon^{(1)}\epsilon^{(2)}}=\left( \sum_{i \neq j} S_{\epsilon_i^{(1)}\epsilon_j^{(2)}} \right)/(N^2-N)$. 
This is interpretable as pertaining to spatial synchrony because it makes comparisons across 
distinct locations. 
It is interpretable as cross-variable synchrony because comparisons are between the two variables, i.e., cross spectra 
between time series components of $\epsilon^{(1)}$ and $\epsilon^{(2)}$ are used. The new index 
takes into account possible time lags. For instance, if both $\epsilon^{(1)}$ and $\epsilon^{(2)}$ show strong,
spatially synchronous, $4$-year-period 
oscillations, but peaks in the $\epsilon^{(2)}$ oscillations consistently lag peaks in the $\epsilon^{(1)}$ oscillations 
by a year, then $\rho_{\epsilon^{(1)}\epsilon^{(2)}}$, which is complex valued, will have high magnitude at
timescale $4$y, equivalent to frequency $f=\frac{1}{4} \text{ y}^{-1}$, and will have  
phase at that frequency equal to $\pi/2$, reflecting the one-year lag.
See Supporting information for detailed examples of cross synchrony.
